---
title: "jjstatsplot Plot Gallery"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE, fig.width=8, fig.height=6)
library(jjstatsplot)
library(ggplot2)
```

This gallery provides a quick visual reference for the main plots you can create with the `ClinicoPath` jamovi module. Each entry includes a brief description of when to use the plot and an example using clinical data.

---

## Histogram

**When to use it:** To explore the distribution of a single continuous variable (e.g., age, lab value).

```{r histogram}
data("breast_cancer_data", package = "ClinicoPath")
jjhistostats(
  data = breast_cancer_data,
  dep = "age",
  title = "Distribution of Patient Age",
  xlab = "Age (years)"
)
```

**jamovi Location:** `JJStatsPlot` -> `Continuous` -> `Histogram`

---

## Box-Violin Plot (Between Groups)

**When to use it:** To compare a continuous variable between two or more independent groups.

```{r box-violin}
data("breast_cancer_data", package = "ClinicoPath")
jjbetweenstats(
  data = breast_cancer_data,
  dep = "age",
  group = "cancer_status",
  mytitle = "Age Distribution by Cancer Status",
  xtitle = "Cancer Status",
  ytitle = "Age (years)"
)
```

**jamovi Location:** `JJStatsPlot` -> `Categorical vs Continuous` -> `Box-Violin Plots (Between Groups)`

---

## Bar Chart

**When to use it:** To show the relationship between two categorical variables.

```{r bar-chart}
data("breast_cancer_data", package = "ClinicoPath")
jjbarstats(
  data = breast_cancer_data,
  dep = "mammography",
  group = "cancer_status"
)
```

**jamovi Location:** `JJStatsPlot` -> `Categorical vs Categorical` -> `Bar Charts`

---

## Scatter Plot

**When to use it:** To explore the relationship and correlation between two continuous variables.

```{r scatter-plot}
data("melanoma", package = "boot")
jjscatterstats(
  data = melanoma,
  dep = "age",
  group = "thickness",
  mytitle = "Correlation between Age and Tumor Thickness",
  xtitle = "Age (years)",
  ytitle = "Tumor Thickness (mm)"
)
```

**jamovi Location:** `JJStatsPlot` -> `Continuous vs Continuous` -> `Scatter Plot`

---

## Correlation Matrix

**When to use it:** To visualize the correlations between multiple continuous variables at once.

```{r corr-matrix}
data("melanoma", package = "boot")
jjcorrmat(
  data = melanoma,
  dep = c("age", "thickness", "time"),
  title = "Correlation Matrix of Melanoma Variables"
)
```

**jamovi Location:** `JJStatsPlot` -> `Continuous vs Continuous` -> `Correlation Matrix`

---

## Within-Subject Plot (Paired Data)

**When to use it:** To compare a continuous variable in the same subjects at two or more time points (e.g., before and after treatment).

```{r within-subject}
# Create simulated paired data
set.seed(123)
long_data <- data.frame(
  patient_id = rep(1:20, 2),
  timepoint = rep(c("Before", "After"), each = 20),
  biomarker = c(rnorm(20, 100, 15), rnorm(20, 80, 15))
)

jjwithinstats(
  data = long_data,
  dep1 = "timepoint",
  dep2 = "biomarker",
  mytitle = "Biomarker Levels Before and After Treatment",
  ytitle = "Biomarker Level"
)
```

**jamovi Location:** `JJStatsPlot` -> `Continuous` -> `Within Subject`

---

## Raincloud Plot

**When to use it:** To show the distribution, individual data points, and summary statistics all in one plot. It is an enhanced version of a violin or box plot.

```{r raincloud-plot}
# Load raincloud data
data("advancedraincloud_data", package = "ClinicoPath")

# Ensure the grouping variable is a factor
advancedraincloud_data$group <- factor(advancedraincloud_data$group)

advancedraincloud(
  data = advancedraincloud_data,
  y_var = "score",
  x_var = "group",
  plot_title = "Score Distribution by Group with Raincloud Plot"
)

```

**jamovi Location:** `JJStatsPlot` -> `Advanced` -> `Raincloud Plot`
