---
title: "jjscatterstats: Comprehensive Scatter Plot Analysis"
author: "ClinicoPath"
date: "`r Sys.Date()`"
format: html
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction to jjscatterstats

The `jjscatterstats` function is a powerful wrapper around `ggstatsplot::ggscatterstats()` and `ggstatsplot::grouped_ggscatterstats()` that provides comprehensive scatter plot analysis with statistical testing capabilities. This function is particularly useful for exploring relationships between continuous variables in clinical and research settings.

## Key Features

- **Statistical Analysis**: Automatic correlation analysis with multiple statistical approaches
- **Flexible Grouping**: Support for grouped analysis across categorical variables
- **Multiple Statistical Types**: Parametric, non-parametric, robust, and Bayesian statistics
- **Customizable Visualization**: Extensive theming and labeling options
- **Performance Optimized**: Enhanced caching and data preparation for faster rendering

## Loading Required Libraries

```{r setup}
library(jjstatsplot)
library(ggplot2)
library(dplyr)

# Load example datasets
data(iris)
data(mtcars)
```

# Basic Scatter Plot Analysis

## Simple Correlation Analysis

Let's start with a basic scatter plot analyzing the relationship between sepal length and petal length in the iris dataset:

```{r basic-example}
# Basic scatter plot with parametric correlation
result_basic <- jjscatterstats(
  data = iris,
  dep = "Sepal.Length",
  group = "Petal.Length",
  typestatistics = "parametric",
  mytitle = "Sepal Length vs Petal Length",
  xtitle = "Sepal Length (cm)",
  ytitle = "Petal Length (cm)"
)

print(result_basic)
```

## Different Statistical Approaches

### Non-parametric Analysis

```{r nonparametric}
# Non-parametric correlation (Spearman's rho)
result_nonparam <- jjscatterstats(
  data = iris,
  dep = "Sepal.Length",
  group = "Petal.Width",
  typestatistics = "nonparametric",
  mytitle = "Non-parametric Correlation Analysis"
)

print(result_nonparam)
```

### Robust Statistics

```{r robust}
# Robust correlation analysis
result_robust <- jjscatterstats(
  data = iris,
  dep = "Sepal.Width",
  group = "Petal.Length",
  typestatistics = "robust",
  mytitle = "Robust Correlation Analysis"
)

print(result_robust)
```

### Bayesian Analysis

```{r bayesian}
# Bayesian correlation analysis
result_bayes <- jjscatterstats(
  data = iris,
  dep = "Sepal.Length",
  group = "Sepal.Width",
  typestatistics = "bayes",
  mytitle = "Bayesian Correlation Analysis"
)

print(result_bayes)
```

# Grouped Analysis

## Scatter Plot by Species

One of the most powerful features is the ability to create grouped scatter plots:

```{r grouped-analysis}
# Grouped scatter plot by species
result_grouped <- jjscatterstats(
  data = iris,
  dep = "Sepal.Length",
  group = "Petal.Length",
  grvar = "Species",
  typestatistics = "parametric",
  mytitle = "Correlation Analysis by Species"
)

print(result_grouped)
```

## Multiple Group Analysis with mtcars

```{r mtcars-grouped}
# Prepare mtcars data
mtcars_modified <- mtcars %>%
  mutate(
    transmission = factor(am, levels = c(0, 1), labels = c("Automatic", "Manual")),
    cylinders = factor(cyl)
  )

# Grouped analysis of mpg vs horsepower by transmission type
result_mtcars <- jjscatterstats(
  data = mtcars_modified,
  dep = "hp",
  group = "mpg",
  grvar = "transmission",
  typestatistics = "parametric",
  mytitle = "MPG vs Horsepower by Transmission Type",
  xtitle = "Horsepower",
  ytitle = "Miles per Gallon"
)

print(result_mtcars)
```

# Clinical Research Applications

## Using Clinical Test Data

```{r clinical-example, eval=FALSE}
# Example with clinical data (if available)
# This demonstrates typical clinical research scenarios

# Load clinical test data
data(jjscatterstats_clinical)

# Analyze tumor size vs Ki67 percentage
clinical_result <- jjscatterstats(
  data = jjscatterstats_clinical,
  dep = "tumor_size_mm",
  group = "ki67_percentage",
  grvar = "stage",
  typestatistics = "parametric",
  mytitle = "Tumor Size vs Ki67 by Cancer Stage",
  xtitle = "Tumor Size (mm)",
  ytitle = "Ki67 Percentage (%)"
)

print(clinical_result)
```

## Biomarker Correlation Analysis

```{r biomarker-example, eval=FALSE}
# Biomarker correlation analysis
biomarker_result <- jjscatterstats(
  data = jjscatterstats_clinical,
  dep = "mutation_count",
  group = "survival_months",
  grvar = "histology",
  typestatistics = "nonparametric",
  mytitle = "Mutation Count vs Survival by Histology",
  xtitle = "Mutation Count",
  ytitle = "Survival (months)"
)

print(biomarker_result)
```

# Advanced Customization

## Theme Customization

```{r theme-custom}
# Using original ggstatsplot theme
result_original_theme <- jjscatterstats(
  data = iris,
  dep = "Sepal.Length",
  group = "Petal.Length",
  originaltheme = TRUE,
  mytitle = "Original ggstatsplot Theme"
)

print(result_original_theme)
```

## Custom Labels and Titles

```{r custom-labels}
# Comprehensive labeling
result_labeled <- jjscatterstats(
  data = mtcars,
  dep = "wt",
  group = "mpg",
  typestatistics = "parametric",
  mytitle = "Vehicle Weight vs Fuel Efficiency",
  xtitle = "Weight (1000 lbs)",
  ytitle = "Miles per Gallon",
  resultssubtitle = TRUE
)

print(result_labeled)
```

## Controlling Statistical Results Display

```{r results-control}
# Hide statistical results subtitle
result_no_stats <- jjscatterstats(
  data = iris,
  dep = "Sepal.Width",
  group = "Petal.Width",
  typestatistics = "parametric",
  mytitle = "Clean Plot Without Statistics",
  resultssubtitle = FALSE
)

print(result_no_stats)
```

# Performance Considerations

## Large Dataset Handling

The function includes several performance optimizations:

```{r performance-demo, eval=FALSE}
# Performance test with larger dataset
data(jjscatterstats_performance)

# This should render efficiently due to optimizations
start_time <- Sys.time()
performance_result <- jjscatterstats(
  data = jjscatterstats_performance,
  dep = "measurement_1",
  group = "measurement_2",
  grvar = "lab_id",
  typestatistics = "parametric",
  mytitle = "Performance Test - Large Dataset"
)
end_time <- Sys.time()

cat("Rendering time:", difftime(end_time, start_time, units = "secs"), "seconds\n")
print(performance_result)
```

## Data Caching and Optimization

The function implements several performance enhancements:

1. **Data Preparation Caching**: Processed data is cached to avoid recomputation
2. **Option Preprocessing**: Common option processing is done once and cached
3. **Hash-based Change Detection**: Only reprocesses data when inputs change
4. **Efficient Memory Usage**: Minimizes data copying and transformation overhead

# Edge Cases and Data Validation

## Handling Missing Values

```{r missing-values}
# Create data with missing values
iris_with_na <- iris
iris_with_na[1:10, "Sepal.Length"] <- NA
iris_with_na[15:20, "Petal.Length"] <- NA

# Function automatically handles missing values
result_na <- jjscatterstats(
  data = iris_with_na,
  dep = "Sepal.Length",
  group = "Petal.Length",
  typestatistics = "parametric",
  mytitle = "Handling Missing Values"
)

print(result_na)
```

## Input Validation

```{r validation-demo, error=TRUE}
# Example of error handling
try({
  result_error <- jjscatterstats(
    data = data.frame(),  # Empty dataset
    dep = "nonexistent",
    group = "variable",
    typestatistics = "parametric"
  )
})
```

# Best Practices and Recommendations

## Statistical Type Selection

- **Parametric**: Use when data is normally distributed (Pearson correlation)
- **Non-parametric**: Use for non-normal data or ordinal variables (Spearman correlation)
- **Robust**: Use when data has outliers (percentage bend correlation)
- **Bayesian**: Use for incorporating prior knowledge or uncertainty quantification

## Visualization Guidelines

1. **Choose appropriate variables**: Both x and y should be continuous
2. **Consider grouping**: Use grouping variables to reveal subgroup patterns
3. **Label clearly**: Always provide meaningful titles and axis labels
4. **Statistical reporting**: Include statistical results unless specifically hiding them

## Performance Tips

1. **Large datasets**: The function is optimized for datasets up to several thousand rows
2. **Multiple analyses**: Reuse data objects when possible to benefit from caching
3. **Grouping variables**: Limit grouping variables to reasonable numbers of levels (< 10)

# Troubleshooting Common Issues

## Variable Type Issues

```{r variable-types}
# Ensure variables are numeric
str(iris[c("Sepal.Length", "Petal.Length")])

# The function automatically converts to numeric, but it's good practice to check
```

## Memory and Performance

```{r memory-tips, eval=FALSE}
# For very large datasets, consider:
# 1. Sampling your data first
set.seed(123)
sampled_data <- iris[sample(nrow(iris), 100), ]

result_sampled <- jjscatterstats(
  data = sampled_data,
  dep = "Sepal.Length",
  group = "Petal.Length",
  typestatistics = "parametric"
)

# 2. Using simpler statistical methods for initial exploration
result_simple <- jjscatterstats(
  data = iris,
  dep = "Sepal.Length",
  group = "Petal.Length",
  typestatistics = "parametric",
  resultssubtitle = FALSE  # Faster rendering
)
```

# Summary

The `jjscatterstats` function provides a comprehensive solution for scatter plot analysis with:

- Multiple statistical testing approaches
- Flexible grouping capabilities
- Performance optimizations for larger datasets
- Extensive customization options
- Robust error handling and validation

This makes it an excellent choice for exploratory data analysis, clinical research, and statistical reporting in the jamovi environment.

## Function Reference

For complete parameter documentation, see:

- [ggstatsplot::ggscatterstats](https://indrajeetpatil.github.io/ggstatsplot/reference/ggscatterstats.html)
- [ggstatsplot::grouped_ggscatterstats](https://indrajeetpatil.github.io/ggstatsplot/reference/grouped_ggscatterstats.html)

```{r session-info}
# Session information
sessionInfo()
```
